---
title: "ggplot2 Tutorial for Bioinformatics"
author: "Zorin Gao <bioinfo.gaoi@gmail.com.com>"
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: true
  pdf:
    pdf-engine: xelatex
    mainfont: "Noto Sans CJK SC"  # 或者使用你系统里有的字体，如 "WenQuanYi Micro Hei" 没有这行，无法显示中文  
execute:
  echo: true
  warning: false
  message: false
---

# ggplot2 Tutorial for Bioinformatics Analysis

针对生物信息学研究（特别是你正在进行的 RNA-seq 和单细胞分析），ggplot2 不仅仅是一个绘图工具，它是一套基于"图形语法"（Grammar of Graphics）的逻辑系统。

以下是一份为你定制的高度详细教程，涵盖了从基础逻辑到生信实战进阶的内容。

## 1. ggplot2 的核心哲学：图层 (Layers)

ggplot2 的绘图逻辑就像在透明胶片上叠画：

- **数据层 (Data)**: 你要画什么？
- **映射层 (Aesthetics)**: 数据中的变量对应到画面的什么位置、颜色、形状？ ========== 最关键！
- **几何对象层 (Geometries)**: 用点、线还是柱状图展示？
- **统计转换层 (Stats)**: 是否需要计算平均值、分位数？
- **标尺控制层 (Scales)**: 颜色用什么色盘？轴的范围是多少？
- **分面层 (Facets)**: 是否要按组分拆成小图？
- **主题层 (Themes)**: 背景、字体、网格线等外观设置。

## 2. 基础语法模板

```{r, run=FALSE}
library(ggplot2)

```

## 3. 核心要素详解

### A. 审美映射 (Aesthetics - aes)

`aes()` 负责将数据列映射到视觉属性。

- `x`, `y`: 坐标轴。
- `color`: 边框或点的颜色。
- `fill`: 填充颜色（如柱状图、箱线图内部）。
- `size`: 大小。
- `shape`: 形状。
- `alpha`: 透明度。

### B. 几何对象 (Geoms)

如果你希望表格有标题或引用编号，可以在表格下方紧接着添加：

| 几何对象 (Geom) | 数据要求 | 生信典型应用场景 | 关键参数建议 |
| :--- | :--- | :--- | :--- |
| **`geom_point()`** | 连续型 x, y | PCA 主成分分析图、火山图 (Volcano Plot) | `alpha=0.5`, `size` |
| **`geom_tile()`** | 离散型 x, y | 差异表达基因热图 (Heatmap) | `aes(fill = value)` |
| **`geom_boxplot()`** | 离散型 x, 连续型 y | 比较不同组别间的基因表达量分布 | `outlier.shape = NA` |
| **`geom_violin()`** | 离散型 x, 连续型 y | 单细胞测序中 Cluster 基因分布 | `trim = FALSE` |
| **`geom_bar()`** | 离散型 x | GO/KEGG 富集分析条形图 | `stat = "identity"` |
| **`geom_density2d()`** | 连续型 x, y | **[Wildcard]** 大样本量单细胞轨迹密度分布 | `color = "blue"` |
| **`geom_jitter()`** | 离散型 x, 连续型 y | 叠加原始数据点 | `width = 0.2`, `alpha = 0.3` |


## 4. Practical Examples

### Example 1: Basic Scatter Plot (PCA-like)
```{r}
# Load required library
library(ggplot2)

# Create sample data similar to PCA results
set.seed(123)
pca_data <- data.frame(
  PC1 = rnorm(100),
  PC2 = rnorm(100),
  group = factor(rep(c("Control", "Treatment"), each = 50))
)

print(pca_data |> head())
print(pca_data |> dim())

# Basic scatter plot
# ggplot(pca_data, aes(x = PC1, y = PC2, color = group)) +
#   geom_point(alpha = 0.7, size = 3) +
#   labs(title = "PCA Plot", x = "PC1", y = "PC2") +
#   theme_minimal()  # +
#   # theme(
#   #   plot.title = element_text(hjust = 0.5, size = 18)
#   # )
# # Use theme() to center the title and increase its size:

ggplot(pca_data, aes(x = PC1, y = PC2, color = group)) +
  geom_point(alpha = 0.7, size = 3) +
  labs(title = "PCA Plot", x = "PC1", y = "PC2") +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 18,
      face = "bold",
      family = "serif"
    ),
    axis.line = element_line(color = "black", linewidth = 1),
    axis.title = element_text(size = 10, face = "bold"), # Axis 标注 单独调axis.title.x、axis.title.y 分开控制
    axis.text = element_text(size = 14, color = "black"),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    legend.background = element_rect(
      color = "black",
      linewidth = 0.8,
      fill = "white"
    )
  )




```

### Example 2: Volcano Plot
```{r}
# Create volcano plot data
volcano_data <- data.frame(
  log2FC = rnorm(1000, mean = 0, sd = 1),
  pvalue = runif(1000, min = 1e-10, max = 1)
)
volcano_data$neg_log10_pval <- -log10(volcano_data$pvalue)
volcano_data$significant <- ifelse(volcano_data$neg_log10_pval > 2 & abs(volcano_data$log2FC) > 1, "Significant", "Not Significant")

# Volcano plot
ggplot(volcano_data, aes(x = log2FC, y = neg_log10_pval, color = significant)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("Significant" = "red", "Not Significant" = "black")) +
  labs(title = "Volcano Plot", x = "Log2 Fold Change", y = "-Log10(P-value)") +
  theme_minimal() +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "gray") +
  geom_hline(yintercept = 2, linetype = "dashed", color = "gray")
```

### Example 3: Box Plot with Jitter Points
```{r}
# Gene expression data across groups
gene_expr <- data.frame(
  expression = c(rnorm(30, mean = 5), rnorm(30, mean = 7)),
  group = factor(rep(c("Control", "Treatment"), each = 30))
)

# Box plot with jitter points
ggplot(gene_expr, aes(x = group, y = expression, fill = group)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) +
  scale_fill_manual(values = c("Control" = "#1f77b4", "Treatment" = "#ff7f0e")) +
  labs(title = "Gene Expression by Group", x = "Group", y = "Expression Level") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
#| label: setup
#| include: false

# Load required libraries
library(ggplot2)
library(dplyr)
```

# Introduction to ggplot2

ggplot2 is a powerful and flexible R package for creating elegant data visualizations based on the Grammar of Graphics. This tutorial will guide you through the fundamentals of ggplot2, from basic plots to more advanced customizations.

## The Grammar of Graphics

The Grammar of Graphics, developed by Leland Wilkinson, provides a structured approach to describing and building statistical graphics. In ggplot2, every plot is composed of:

1. **Data**: The dataset being visualized
2. **Aesthetics**: How variables map to visual properties (x, y, color, size, etc.)
3. **Geoms**: Geometric objects that represent the data (points, lines, bars, etc.)
4. **Stats**: Statistical transformations (binning, smoothing, etc.)
5. **Scales**: How data values map to visual properties
6. **Coordinate system**: The system used to position elements
7. **Facets**: How to break data into subsets for separate panels
8. **Themes**: Non-data ink (fonts, backgrounds, gridlines, etc.)

## Basic Structure of ggplot2

The basic structure of a ggplot2 plot follows this pattern:

```{r, run=FALSE}
# Basic ggplot2 structure
# ggplot(data = <DATA>) +
#   <GEOM_FUNCTION>(
#     mapping = aes(<MAPPINGS>)
#   )
```

Let's start with a simple example using the built-in `mtcars` dataset:

```{r}
# Simple scatter plot
ggplot(data = mtcars) +
  geom_point(mapping = aes(x = wt, y = mpg))
```

## Aesthetics Mapping

Aesthetics define how variables in your data are mapped to visual properties. Common aesthetics include:

- `x` and `y`: Position on x and y axes
- `color`/`colour`: Color of points, lines, or areas
- `fill`: Fill color for areas
- `size`: Size of points or areas
- `alpha`: Transparency (0 = fully transparent, 1 = fully opaque)
- `shape`: Shape of points
- `linetype`: Type of line (solid, dashed, etc.)

### Example: Adding Color Based on Cylinders

```{r}
# Scatter plot with color mapped to number of cylinders
ggplot(data = mtcars) +
  geom_point(mapping = aes(x = wt, y = mpg, color = factor(cyl)))
```

Note that we use `factor(cyl)` to treat cylinder count as a categorical variable rather than continuous.

## Common Geoms

Here are some of the most commonly used geometric objects:

### Points (`geom_point`)
```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point()
```

### Lines (`geom_line`)
```{r}
# Create time series data
time_data <- data.frame(
  time = 1:10,
  value = cumsum(rnorm(10))
)

ggplot(time_data, aes(x = time, y = value)) +
  geom_line()
```

### Bars (`geom_bar` and `geom_col`)
```{r}
# geom_bar counts occurrences
ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar()

# geom_col uses values as heights
bar_data <- data.frame(
  category = c("A", "B", "C"),
  value = c(10, 15, 7)
)

ggplot(bar_data, aes(x = category, y = value)) +
  geom_col()
```

### Histograms (`geom_histogram`)
```{r}
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(bins = 10)
```

### Boxplots (`geom_boxplot`)
```{r}
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot()
```

## Layers and Multiple Geoms

You can add multiple layers to a single plot:

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "red")
```

## Faceting

Faceting creates multiple plots based on one or more variables:

### `facet_wrap()`: Wraps facets into a grid
```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  facet_wrap(~ cyl)
```

### `facet_grid()`: Creates a grid of rows and columns
```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  facet_grid(vs ~ am)
```

## Scales and Labels

Customize how data maps to visual properties and add informative labels:

```{r}
ggplot(mtcars, aes(x = wt, y = mpg, color = hp)) +
  geom_point(size = 3) +
  scale_color_gradient(low = "blue", high = "red") +
  labs(
    title = "Fuel Efficiency vs Weight",
    subtitle = "Colored by Horsepower",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon",
    color = "Horsepower"
  )
```

## Themes

Themes control the non-data elements of your plot:

```{r}
# Using built-in themes
p <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point()

# Different theme options
p + theme_bw()
```

```{r}
# Custom theme
p + theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    panel.grid.minor = element_blank()
  ) +
  labs(title = "Custom Styled Plot")
```

## Saving Plots

Save your plots to files:

```{r}
#| eval: false
p <- ggplot(mtcars, aes(x = wt, y = mpg)) + geom_point()

# Save as PNG
ggsave("my_plot.png", plot = p, width = 8, height = 6, dpi = 300)

# Save as PDF
ggsave("my_plot.pdf", plot = p, width = 8, height = 6)
```

## Best Practices

1. **Start simple**: Begin with basic plots and add complexity gradually
2. **Think about your audience**: Choose appropriate colors, labels, and scales
3. **Use meaningful labels**: Always label axes and provide clear titles
4. **Consider colorblindness**: Use color palettes that are accessible to all viewers
5. **Avoid unnecessary elements**: Remove chart junk that doesn't add information

## Advanced Example: Multi-panel Publication Quality Plot

Let's create a more complex, publication-ready plot:

```{r}
# Create sample data
set.seed(123)
sample_data <- data.frame(
  group = rep(c("Control", "Treatment"), each = 50),
  time = rep(1:10, times = 10),
  value = c(rnorm(50, mean = 5), rnorm(50, mean = 7)) + 
          rep(seq(0, 2, length.out = 10), times = 10)
)

# Create the plot
ggplot(sample_data, aes(x = time, y = value, color = group)) +
  geom_line(aes(group = interaction(group, rep(1:5, each = 10))), 
            alpha = 0.3) +
  geom_smooth(se = TRUE, method = "loess") +
  scale_color_manual(values = c("Control" = "gray50", "Treatment" = "steelblue")) +
  labs(
    title = "Treatment Effect Over Time",
    x = "Time (hours)",
    y = "Response Value",
    color = "Group"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "bottom"
  )
```

## Resources for Further Learning

- [ggplot2 official documentation](https://ggplot2.tidyverse.org/)
- [R Graph Gallery](https://r-graph-gallery.com/)
- ["R Graphics Cookbook" by Winston Chang](https://r-graphics.org/)
- [Data Visualization: A Practical Introduction by Kieran Healy](https://socviz.co/)

This tutorial covers the fundamentals of ggplot2. With practice, you'll be able to create sophisticated and publication-quality visualizations for your data analysis projects.